Something about Performance, 2D Component of NEMO. (TBC)

### 2016-01-11 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d paper: fix typos and tweak text. Update details of shallow paper and correct citation of OpenMP

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain.

{\bf Auto-generated BC code by reasoning about meta-data. But then,
  that brings us on to meta-data...}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology(?)}
\label{sec_methodology}

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
3.7~GHz and 10~MB of last-level cache. The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and hardware used in this work. The
  use of a compiler on a given system is indicated by the specification
  of the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that system.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                 & \multicolumn{5}{c|}{Compiler}            \\
\hline
                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
\hline
Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
IBM Power 8      &       &             &         &        &  \\
NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
  Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

@@

\subsubsection{Addition of `safe-address' directives}
\label{sec_safe_address}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the safe\_address
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

 A loop starting at line 99 was not
 vectorized because it contains 
 conditional code which is more
 efficient if executed in scalar mode.
\end{verbatim}
Inserting the Cray `vector always' directive persuaded the compiler
to vectorise the loop:
\begin{verbatim}
  99.    !dir$ vector always
  100.          do ji = 2, M-1, 1
 A loop starting at line 100 was blocked
 with block size 256.

@@

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities ({\it e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were obtained from the
same object as that containing the array to which that kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

@@

The simplest possible OpenMP-parallel implementation consists of
parallelising each loop nest in the PSy layer. This was done by
inserting an OpenMP PARALLEL DO directive before each loop nest so
that the iterations of the outermost or $j$ loop (over the y-dimension
of the model domain) are shared amongst the OpenMP threads. This
leaves the innermost ($i$) loop available for SIMD vectorisation by
the compiler.

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity has a loop-carried dependency
in $j$ which prevented its being executed in parallel. This was
therefore left unchanged and executed on thread 0 only.

@@

Although very simple to implement, the use of separate PARALLEL DO
directives causes the team of OpenMP threads to be repeatedly created
and destroyed. This may be avoided by keeping the thread team in
existence for as long as possible using an OpenMP PARALLEL region.  We
therefore enclosed the whole of the PSy layer (in this code, a single
subroutine) within a single PARALLEL region. The directive preceeding
each loop nest to be parallelised was then changed to an OpenMP DO. We
ensured that the v-Flather loop nest was executed in serial (by the
first thread to encounter it) by enclosing it within an OpenMP SINGLE
section.

@@

Since data arrays are managed within the GOcean Library {\bf (is this
  a good/appropriate name?)} this optimisation can again be
implemented without changing user code (the Application and Kernel
layers).

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the nemolite2d code for the range of
compilers and CPUs considered here. Unlike the Shallow code, the
original version of nemolite2d has not been optimised. Although it is
still a single source file it is, in common with NEMO itself,
logically structured with separate subroutines performing different
parts of the physics within each time step. This structuring and the
heavy use of conditional statements favour the Intel compiler which
significantly out-performs both the Gnu and Cray compilers. Only when
the problem size falls out of cache does the performance gap begin to
narrow. The reason for the performance deficit of the Cray-compiled
binary comes down to SIMD vectorisation, an issue that we explore
below.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Performance (millions of points updated per second) for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.62 & 2.39 & 42.22 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=90mm]{opt_stages_256}
\caption{Performance of the \psykal version of nemolite2d for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  nemolite2d for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
nemolite2d code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
e.g. the construction of field objects.

@@

In restricting ourselves to a single socket, we are keeping all
threads within a single NUMA region. It is therefore surprising that
implementing a `first-touch' policy has any effect and yet, for the
Gnu- and Intel-compiled binaries, it appears to improve performance
when hyperthreading is employed to run on 24 threads (green lines and
diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
~\ref{FIG_intel_omp_256}). {\bf Can we explain this?}

@@

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
configuration we ensured that threads were evenly shared over the two
sockets. The performance obtained for the $256^2$ case with the `early
SINGLE' version of the code is compared with the single-socket
performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
available memory bandwidth in this way has little effect on
performance - the two-socket performance figures track those from a
single socket very closely. The only realy difference in performance
is at 24 threads where, in addition to the difference in available
memory bandwidth, the single-socket configuration is using
hyperthreading while the two-socket case is not.  {\bf Have query in
  with Archer helpdesk on behaviour seen with Cray compiler here. Also
  would like to run 512 case on 2 sockets to see how it goes out to 48
  threads.}

>>>>

As with any full oceanographic model, boundary conditions must be
applied at the edges of the model domain. Since nemolite2d applies
external boundary conditions ({\it e.g.} barotropic forcing), this is
done via user-supplied kernels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology}
\label{sec_methodology}

Our aim in this work is to achieve portable performance. Consequently,
we have performed tests with the hardware and compilers listed in
Table~\ref{TABLE_compilers}. Where a compiler is available on a given
piece of hardware, the version number used in this work is specified.
%The Intel Haswell-based system (Xeon E5-1620 v2) has a clock speed of
%3.7~GHz and 10~MB of last-level cache. 
The Intel Ivy Bridge-based
system (Xeon E5-2697) has a clock speed of 2.7~GHz and a last-level
cache of 30~MB.

@@

% An example of a floating table. Note that, for IEEE style tables, the 
% \caption command should come BEFORE the table. Table text will default to
% \footnotesize as IEEE normally uses this smaller font for tables.
%
\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The matrix of compilers and hardware used in this work. The
  use of a compiler on a given system is indicated by the specification
  of the version of the compiler in the relevent element. No entry
  indicates that a compiler was not available/used on that system.}
\label{TABLE_compilers}
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                 & \multicolumn{5}{c|}{Compiler}            \\
\hline
                 & Gnu   & Intel       & Cray    & IBM    &  PGI \\
\hline
%Intel Haswell    & 4.8.2 & 14.0.0      &         &        &  \\
Intel Ivy Bridge & 4.9.1 & 15.0.0.090  & 8.3.3   &        &  \\
IBM Power 8      &       &             &         &        &  \\
NVIDIA Tesla K40 &       &             &         &        &  {\bf Jeremy} \\
\hline
\end{tabular}
\end{table}

@@

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d on the Intel Ivy Bridge CPU (for 2000 time-steps of the
  $128^{2}$ domain and the Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c|}{Vanilla \psykal} &
\multicolumn{2}{c}{Final \psykal} \\
        & Time (s) & \%  & Time (s) & \%  & Time(s) & \% \\
\hline
Momentum    & 1.98  & 72.6  & 4.05  & 79.5 & 2.09 & 75.3 \\
Time-update & 0.40  & 14.6  & 0.41  & 8.1  & 0.29 & 10.6 \\
BCs         & 0.25  & 9.1   & 0.29  & 5.7  & 0.27 & 9.9  \\
Continuity  & 0.10  & 3.7   & 0.33  & 6.6  & 0.11 & 4.1  \\
\hline
\end{tabular}
\end{table}

@@

\subsubsection{Addition of {\tt safe\_address} directives}
\label{sec_safe_address}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the {\tt safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

 A loop starting at line 99 was not
 vectorized because it contains 
 conditional code which is more
 efficient if executed in scalar mode.
\end{verbatim}
Inserting the Cray {\tt vector always} directive persuaded the compiler
to vectorise the loop:
\begin{verbatim}
  99.    !dir$ vector always
  100.          do ji = 2, M-1, 1
 A loop starting at line 100 was blocked
 with block size 256.

@@

Comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to limitations in the compiler's analysis of
the way components of Fortran derived types were being used. Each
GOcean field object, in addition to the array holding the local
section of the field, contains a pointer to a GOcean grid object. If a
kernel requires grid-related quantities ({\it e.g.} the grid spacing)
then these are obtained by passing it a reference to the appropriate
array within the grid object. Although these grid-related quantities
are read-only within a compute kernel, if they were referenced from the
same field object as that containing an array to which the kernel writes
then the Intel compiler identified a dependency preventing
vectorisation.  This limitation was simply removed by ensuring that
all read-only quantities were accessed via field objects that were
themselves read-only for the kernel at hand.  For instance, the call
to the continuity kernel, which confused the Intel compiler, originally
looked like this:
\begin{verbatim}
do jj = 2, N, 1
 do ji = 2, M, 1

@@

The simplest possible OpenMP-parallel implementation consists of
parallelising each loop nest in the PSy layer. This was done by
inserting an OpenMP PARALLEL DO directive before each loop nest so
that the iterations of the outermost or $j$ loop (over the y-dimension
of the model domain) are shared out amongst the OpenMP threads. This
leaves the innermost ($i$) loop available for SIMD vectorisation by
the compiler.

The loop nest dealing with the application of the Flather boundary
condition to the y-component of velocity has a loop-carried dependency
in $j$ which appears to prevent its being executed in parallel. This was
therefore left unchanged and executed on thread 0 only.

@@

Although very simple to implement, the use of separate PARALLEL DO
directives results in a lot of thread synchronisation and can also
cause the team of OpenMP threads to be repeatedly created and
destroyed. This may be avoided by keeping the thread team in existence
for as long as possible using an OpenMP PARALLEL region.  We therefore
enclosed the whole of the PSy layer (in this code, a single
subroutine) within a single PARALLEL region. The directive preceeding
each loop nest to be parallelised was then changed to an OpenMP DO. We
ensured that the v-Flather loop nest was executed in serial (by the
first thread to encounter it) by enclosing it within an OpenMP SINGLE
section.

@@

Since data arrays are managed within the GOcean Library this
optimisation can again be implemented without changing user code ({\it
  i.e.} the Application and Kernel layers).

@@

In Figure~\ref{FIG_orig_perf_summary} we plot the serial performance
of the original version of the nemolite2d code for the range of
compilers and CPUs considered here. Unlike the Shallow
code~\cite{shallow_psykal}, the original version of nemolite2d has not
been optimised. Although it is still a single source file it is, in
common with NEMO itself, logically structured with separate
subroutines performing different parts of the physics within each time
step. This structuring and the heavy use of conditional statements
favour the Intel compiler which significantly out-performs both the
Gnu and Cray compilers. Only when the problem size falls out of cache
does the performance gap begin to narrow. The reason for the
performance deficit of the Cray-compiled binary comes down to (a lack
of) SIMD vectorisation, an issue that we explore below.

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Performance (millions of points updated per second) for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
CPU:                & \multicolumn{3}{c}{Ivy Bridge}  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla \psykal  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best {\it c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=93mm]{opt_stages_256}
\caption{Performance of the \psykal version of nemolite2d for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  nemolite2d for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

We now turn to transformations related to parallelisation of the
nemolite2d code; the introduction of OpenMP and OpenACC directives. In
keeping with the \psykal approach we do not modify either the
Algorithm- or Kernel-layer code.  Any code changes are restricted to
either the PSy (middle) layer or the underlying library that manages
{\it e.g.} the construction of field objects.

@@

In restricting ourselves to a single socket, we are keeping all
threads within a single NUMA region. It is therefore surprising that
implementing a `first-touch' policy has any effect and yet, for the
Gnu- and Intel-compiled binaries, it appears to improve performance
when hyperthreading is employed to run on 24 threads (green lines and
diamond symbols in Figures~\ref{FIG_gnu_omp_256} and
~\ref{FIG_intel_omp_256}). {\bf Can we explain this? Perhaps with HW
  counters?}

@@

Since NEMO and similar finite-difference codes tend to be
memory-bandwidth bound, we checked the sensitivity of our performance
results to this quantity by benchmarking using two sockets of Intel
Ivy Bridge ({\it i.e.} using a complete node of Archer). For this
configuration we ensured that threads were evenly shared over the two
sockets. The performance obtained for the $256^2$ case with the `early
SINGLE' version of the code is compared with the single-socket
performance in Figure~\ref{FIG_omp_2socks}. Surprisingly, doubling the
available memory bandwidth in this way has little effect on
performance - the two-socket performance figures track those from a
single socket very closely. The only significant difference in performance
is at 24 threads where, in addition to the difference in available
memory bandwidth, the single-socket configuration is using
hyperthreading while the two-socket case is not.  {\bf Have query in
  with Archer helpdesk on behaviour seen with Cray compiler here. Also
  would like to run 512 case on 2 sockets to see how it goes out to 48
  threads.}

{\bf Add roofline model to aid understanding of this result.}
